What is an Ancestral Recombination Graph?

### 2022-03-09 paper.tex
Another pass on SPRs

\section*{ARGs and SPRs}
\citet{wiuf1999ancestry,wiuf1999recombination} recast
the coalescent with recombination
as a process along the genome rather than backwards in time.
In this formulation we begin at one end of a chromosome by
sampling a tree under the standard coalescent, and then reason
about how this tree is modified by recombination events
as we move along the genome. Although this algorithm is
significantly more complicated to simulate than Hudson's
backwards in time approach, it has led to substantial
insights into the process of recombination itself
and to the sequentially Markov Coalescent (SMC)
approximations~\citep{mcvean2005approximating,marjoram2006fast}
which are central to many modern methods.

Under the coalescent with recombination (or SMC) each local
tree along the genome is transformed into the next by
a ``subtree prune-and-regraft'' (SPR)
% FIXME there should be earlier refs for SPRs, right?
operation~\citep{song2003on,song2006properties}.
The number of SPRs required to transform one leaf-labelled tree
into another is known as the SPR distance [cite]
and is computable in O(??) time.

@@

The equivalence between an ARG and a set of local trees
separated by SPRs is often mentioned [e.g. x y z], but
it is not obvious how to produce an ARG given (say) a
collection of newick trees. If the internal nodes are
labelled, then we can simply input the edges for each
tree in turn, which can easily be ``squashed'' down
to a minimal representation. However, internal nodes
are rarely labelled in newick trees, and therefore
the process of contructing a graph is much harder.
In princple, we can contruct an EARG by reasoning
about the SPRs that must has occured between two
adjacent trees and postulating the recombination
nodes that must have occured. However, this process
is computationally expensive [How expensive? is this
what Espalier does?] Additionally, unless some extra
information is provided the time of the recombination
will not be known. Fundamentally, we must include the
unary nodes corresponding to recombination parents
in the trees if we wish to recover an ARG from them.

>>>>

\section*{ARGs and SPRs}
\citet{wiuf1999ancestry,wiuf1999recombination} recast
the coalescent with recombination
as a stocastic process along the genome rather than backwards in time.
In this formulation we begin at one end of a chromosome by
sampling a tree under the standard coalescent, and then reason
about how this tree is modified by recombination events
as we move along the genome. Although this algorithm is
significantly more complicated to simulate than Hudson's
backwards in time approach, it has led to substantial
insights into the process of recombination itself
and to the sequentially Markov Coalescent (SMC)
approximations~\citep{mcvean2005approximating,marjoram2006fast}
which are central to many modern methods.

Under the coalescent with recombination (or SMC) each local
tree along the genome is transformed into the next by
a ``subtree prune-and-regraft'' (SPR)
% Hein et al call it the "subtree transfer" operation but it's
% the same thing.
operation~\citep{hein1990reconstructing,song2003on,song2006properties}.
The number of SPRs required to transform one leaf-labelled tree
into another is known as the SPR distance
% Hein 1996 say that SPR distance is NP-hard, but Allen 2001
% say this is wrong. But then Bordewich went ahead and said
% yes, actually it is NP hard.
and is NP-hard to
compute~\citep{hein1996complexity,allen2001subtree,bordewich2005computational}.

@@

The equivalence between an ARG and a set of local trees
separated by SPRs is often mentioned [e.g. x y z], but
it is important to note that while extracting local
trees from an ARG is straightforward, the converse
process of reconstructing precisely the same ARG from
these trees is more problematic. If the internal
nodes are consistently labelled and unary nodes are
included in the local trees (e.g., Fig~\ref{fig-simplification})
then it is easy to exactly reconstruct the ARG.
Otherwise, the problem is much more difficult and it
may not be possible to uniquely reconstruct the original ARG.
Without labels for the internal nodes, we are forced to
to reason about which subtrees remain the same and which
differ due to the SPR separating adjacent trees in order to
assign node identity. Without the internal unary nodes marking
recombinations, we must infer their existence. In
general, there will not be a unique solution to this problem,
and so there can be many possible ARGs compatible with
a given list of
% want to say "standard phylogenetic" trees here or something, but
% and they don't have to be binary (just arity > 1), but thought
% this would be familiar to readers?
leaf-labelled binary trees.
